Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-27 pharmaco.Rmd
separate script for paper 1 tables

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")

@@

_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as increased risk (listed conditions). We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.

_Methods_ -- Cases of severe COVID-19 were ascertained as all those who tested positive for SARS-CoV-2 nucleic acid in the national database and entered a critical care unit or died within 28 days of testing.  From the national population register ten controls per case were selected, matched for sex, age and general practice. International Classification of Disease (ICD-10) codes were extracted from the past three years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months. Rate ratios were estimated by conditional logistic regression.

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus no dispensed prescriptions was `r table.numdrugsgr.listed.any[6, 4]`. In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.  The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`. 
A backwards elimination procedure identified proton pump inhibitors and opiate analgesics as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of these associations showed dose-response relationships that were strongest in younger age groups and reduced only slightly by adjusting for prespecified covariates.  Stratification by time window of exposure showed that the rate ratios were higher for those with prescriptions dispensed only in the last 120 days than for those with prescriptions dispensed only in an earlier time window.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.  This requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of opiates and proton pump inhibitors should be reinforced.

@@

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.   This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.

# Methods
# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.

For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry. Diagnoses of diabetes were extracted from linkage to the national diabetes register.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.  Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).  The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items, including compound analgesics.

As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).

@@

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).

Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.

To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  
A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and nonopioid or compound analgesics.

The associations with these drug classes were examined in more depth.  For studies of these specific exposures the observations were not restricted to exclude those with listed conditions.  Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.  We used several approaches:

@@

* Adjusting for prespecified covariates: for proton pump inhibitors these covariates were specified as hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  For opiate use any history of neoplasm was included as a covariate.  Care home residence and SIMD quintile were included as covariates for both these drug classes.

* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing

## Associations of severe COVID-19 with opiate use
Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship.  With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.

Table \ref{tab:timewindow.opiate shows that the rate ratio associated with exposure in the most recent 120-day time window was higher than that associated with exposure in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations of severe COVID-19 with proton pump inhibitor use
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].

Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.

The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019]. These indicators of overprescribing include proton pump inhibitors and analgesics, identifed by backwards elimination as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy. 
Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation. The time of exposure to a drug does not necessarily correspond to the time intervals for which the prescription was issued.   Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.

For both opiates and proton pump inhibitors, four criteria provide modest evidence favouring causal explanations over confounding.  There are strong dose-response relationships of COVID-19 to dispensed average daily dose.  Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly. The effect sizes are larger in younger individuals, in whom the prevalence of comorbid conditions and frailty is lower than in older individuals.  Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.  We emphasize however that this   evidence is not conclusive. It may be possible to resolve this through studies in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.  In Scotland proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018], but overprescribing of opiates is still relatively uncommon.

The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration before 2012 [@edelman_association_2019].  As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection.

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest. We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

## Methodological strengths and weaknesses
Although we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints, we have diagnoses from hospital discharge records that have been coded to ICD-10 by trained coders. A strength of this study is that it is based on dispensed rather than issued prescriptions.

@@

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.  The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [github](https://github.com/pmckeigue/covid-scotland_public).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

@@

```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood when that drug class is dropped from a model with care home as covariate"}

@@

```{r numdrugs.cardio, echo=FALSE, warning=FALSE, message=FALSE}

colnames(table.jointcardiovasc)[c(1, 4:5)] <- c("Drug category", "Odds ratio (95\\% CI)",
                                                "$p$-value")

knitr::kable(table.jointcardiovasc,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.cardio",
             row.names=FALSE,
             caption="Multivariate regression of severe COVID-19 on number of cardiovascular and number of noncardiovascular drug classes in dispensed prescriptions, among those without listed conditions") %>%
    column_spec(column=1, width="2.5cm") %>%
    add_header_above(c(" "=1, "Mean number of drug classes"=2)) %>%
    kable_styling(latex_options="hold_position")

@@

```{r dosegr.opiate, echo=FALSE, warning=FALSE, message=FALSE}

>>>>

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n")

# abstract limit 250 words

@@

_Background_ -- Dispensing of at least one prescription in the past year is a risk factor fo severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.

_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the 4320 cases of severe COVID-19 in Scotland since the start of the epidemic.  Cases and controls were linked to records of hospital discharges in the last three years and dispensed prescriptions issued in primary care since June 2019.

*Results* -- In those without a diagnosed risk condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.numdrugsgr.listed.any[">12", 4]`.  Cardiovascular drugs did not contribute to this association: the largest independent contributors were proton pump inhibitors (PPI) and analgesics.  For both PPI and opiate-containing analgesics, there were dose response relationships to severe COVID-19 that were strongest in those aged under 75 years: rate ratios were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opiate["0-74 years: >50", 4]` for $\ge 50$ MME opiates, The association was stronger with recent than with non-recent exposure.

*Conclusions* -- In Scotland severe COVID-19 is associated with polypharmacy and this is not easily explained by co-morbidity.  Although the evidence for causality of proton pump inhibitors and opiates is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.

@@

# Background
In the initial analysis of REACT-SCOT, a matched case control study of severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.  This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated by public health agencies as conferring vulnerability to COVID-19 (hereafter listed conditions).  The objective of this study was to investigate this association.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.

For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry. Diagnoses of diabetes were extracted from linkage to the national diabetes register.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opiates were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).  Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).  The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items.

As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic obstructive airway disease, chronic kidney disease, disabling neurological disease, liver disease and immunodeficiency or immunosuppression.  ICD-10 diagnostic and BNF drug codes used to derive these conditions are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).  Socioeconomic status was encoded as the quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).

@@

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.  In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions. To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (subparagraph code 0407010).  The items grouped in this subparagraph include opiate-containing compound preparations.  `r paste.colpercent(with(subset(cc.severe, CASE==0), table(nonopioid.analgesic, opioid.analgesic)))[2, 2]` of those with a dispensed prescription of a drug in subparagraph 0407020 (opiate analgesics) also had a prescription in subparagraph 0207010.

To examine the associations with these drug classes in more depth, the total dispensed doses of proton pump inhibitors (as DDDs) of proton pump inhibitors and opiate (as MME, summed over subparagraphs 0407010 and 0407020) were calculated.  Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.  We used several approaches:

@@

* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.

* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing.

## Associations with opiate use
Table \ref{tab:dosegr} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship, strongest in those aged less than 75 years.  With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with opiate dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations with proton pump inhibitor use
Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for COVID-19 increases with the dose of proton pump inhibitor.  The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile and any history of neoplasm.  This dose-response relationship was absent in those aged 75 years and over.  In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 2 DDDs/day MME /day were `r table.dosegr.protonpump.notlisted["2 or more", 4]` and `r table.dosegr.protonpump.notlisted["2 or more", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation. in individuals without conditions designated as conferring high risk.  The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017]. We attempted to investigate how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of specific drugs.  A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.  Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10, and that drug exposure is based on dispensed rather than issued prescriptions.

The drug classes that are strongly associated with severe COVID-19 include proton pump inhibitors and analgesics, widely recognized to be indicators of overprescribing as in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019]. As these are among the most commonly prescribed drug classes, it is to be expected that they are selected by elimination as having the largest independent contributions to the association of COVID-19 with polypharmacy. Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation. Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.

For both opiates and proton pump inhibitors, four criteria provide moderate but not conclusive evidence favouring causal explanations over confounding.  There are strong dose-response relationships of COVID-19 to dispensed average daily dose.  Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly. The effect sizes are larger in younger individuals and when the analysis is restricted to those without any of the designated risk conditions for COVID-19.  Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window. It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.

The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration from 2000-2012 [@edelman_association_2019].  As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase susceptibility to severe infection.

We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

@@

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.  Datasets were de-identified before analysis.  The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

@@

```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood of a conditional logistic regression of severe COVID-19 on care home and polypharmacy as covariates when that drug class is dropped, in  those without listed conditions."}

@@

```{r dosegr, echo=FALSE, warning=FALSE, message=FALSE}

colnames(table.dosegr.protonpump) <- colnames(table.dosegr.opiate)
table.dosegr <- rbind(table.dosegr.opiate, table.dosegr.protonpump)

knitr::kable(table.dosegr,
             escape=FALSE,
             label="dosegr", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average dose", colnames(table.dosegr.opiate)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of opiate (MME) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
	column_spec(column=1, width="3cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 16) %>%
    kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
    kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
    kableExtra::group_rows("Ages 0-74 years", 9, 13) %>%
    kableExtra::group_rows("Ages 75+ years", 14, 18) %>%
    landscape()

@@

```{r timewin, echo=FALSE, warning=FALSE, message=FALSE}

colnames(table.timewindow.protonpump) <- colnames(table.timewindow.opiate)
table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opiate), 2),
                               table.timewindow)

knitr::kable(table.timewindow,
             escape=FALSE,
             label="timewindow", 
             booktabs=TRUE,
             row.names=FALSE, 
             col.names=c("Time window", clean.header(colnames(table.timewindow.opiate)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption=paste0("Associations of severe disease with dispensed prescriptions by ", TW, "-day time window")) %>%
	column_spec(column=1, width="3cm") %>%
	column_spec(column=2:3, width="2cm") %>%
	column_spec(column=4, width="2.5cm") %>%
	column_spec(column=5, width="1.5cm") %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows(group_label="Opiate-containing analgesics", 1, 4) %>%
    kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
```
